---
title: "SI Course Project"
author: "Yigit Ozan Berk"
date: "6/27/2019"
output: pdf_document
---

# Overview
In a few (2-3) sentences explain what is going to be reported on.


# Simulations

Initiation of required packages:
```{r}
library(ggplot2)

```


Simulating 1000 sets of 40 exponential random variables:

```{r}
set.seed(121212)
my_lambda <- 0.2
my_n <- 40
nsim <- 1000

samples <- matrix(nrow = 40, ncol = 1000)
for(i in 1:1000) samples[,i] = rexp(my_n, my_lambda)
```

Means, Standard Deviations, and Variances of simulated 40 exponential random variables:

```{r}
mns <- apply(samples, 2, mean)

sdevs <- apply(samples, 2, sd)

vars <- apply(samples, 2, var)
```

The theoretical mean of the distribution is lambda = 1/0.2 = `1/.2`

The overall average of 1000 simulated 40 exponential random variables are `mean(mns)`


```{r}
hist(mns, main = "Distribution of Averages",
     xlab = "E[X']")
abline(v = 5, col = "magenta", lwd = 2)
abline(v = quantile(mns, c(.025, .975)), col = "lightblue", lwd = 2)
```

The distribution of averages is centered around the theoretical mean as expected; with 95% confidence interval:
```{r}
quantile(mns, c(.025, .975))
```

With a theoretical variance of 25, the distribution of observed variances are as follows

```{r}
hist(vars, main = "Distribution of Variances",
     xlab = "var")
abline(v = 25, col = "magenta", lwd = 2)
abline(v = quantile(vars, c(.025,.975)), col = "lightblue", lwd = 2)
```

```{r}
hist(sdevs, main = "Distribution of St. Deviations",
     xlab = "sd")
abline(v = 5, col = "magenta", lwd = 2)
abline(v = quantile(sdevs, c(.025,.975)), col = "lightblue", lwd = 2)
```



# Sample Mean versus Theoretical Mean


# Sample Variance versus Theoretical Variance


# Distribution


# Part 2 - ToothGrowth Data Analysis

