---
title: 'Statistical Inference : Week 2 Classes'
author: "Yigit Ozan Berk"
date: "6/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 2

Variability
Distributions
Asymptotics
Excercises in Swirl
Quiz

# Variability

Var(X) = E[(X - mu)^2] = E[X^2] - E[X]^2

expected value of [x^2] minus the expected value of X quantity squared

square root of the variance is called standard deviation

E[X] = 3.5 - x is a toss of a die

E[X^2] = 1^2 * 1/6 + 2^2 * 1/6 + ....... + 6^2 * 1/6 = 15.17

Var(X) = E[X^2] - E[X]^2 equals approximately 2.92

the variance of a biased coin is p*(1-p)



population variance and the sample variance is directly proportional

Var(X) = E[(X - mu)^2] = E[X^2] - E[X]^2

Sample variance = S^2

sample variance is the average square of distance of the observed observations minus the sample mean.

the expected value of the sample variance is the population variance


# standard error

we call the standard deviation of a statistic 'a standard error'


#data example

```{r}
library(UsingR)
data(father.son)
x <- father.son$sheight
n <- length(x)
round(c(var(x), var(x)/n, sd(x), sd(x)/sqrt(n)), 2)
```

0.09 is the standard deviation in the distribution of averages of n children's heights

# Summary

the sample variance estimates the population variance

the distribution of the sample variance is centered at what it's estimating

it gets more concentrated around the population variance with larger sample sizes

the variance of the sample mean is the population variance devided by n

# Important distribution types

## Binomial distribution

'bernoulli distribution' - arises as the result of a binary outcome

only takes values 1 or 0 with probability p

variance is p*(1-p)

### binomial trials 

binomial random variable is the total number of heads in a potentially biased coin.

```{r}
choose(8,7) * .5^8 + choose(8,8) * .5^8

pbinom(6, size = 8, prob = .5, lower.tail = FALSE)
```


## normal distribution

gaussian distribution


```{r}
pnorm(1160, mean = 1020, sd = 50, lower.tail = F)

#is the same as

pnorm(2.8, lower.tail = FALSE)

#because (1160-1020)/50 equals 2.8
#1160 is 2.8 standard deviations away from the mean

#when calculating in terms of standard normals, that's x=2.8
#because sd = 1



```

## Poisson distribution

used to model counts

uses for poisson distribution

- modeling count data
- modeling event-time or survival data
- modeling contingency tables
- approximating binomials when n is large and p is small

```{r}
pbinom(2, size = 500, prob = 0.01)

ppois(2, lambda = 500 * 0.01)
```

# Asymptotics


